On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-25 tex/draft/draft4.tex
removed c subscript from Mcrit

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.eps}
%\vspace{-0.5in}
\caption{The critical core mass as a function of semimajor axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is for Solar abundances). The decline in $M_{\rm c,crit}$ with distance is a robust result for standard disk models.  The analytic model, which neglects self-gravity, over-predicts the steepness of the decline.}
\label{fig:Mcvsa}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.eps}
%\vspace{-0.5in}
\caption{Critical core masses vs. distance for standard and reduced (by a factor of ten) dust opacities.  Lower opacities give significantly lower $M\crit$ values.  Atmospheric opacity remains a large uncertainty in core accretion models.}
\label{fig:Mcritopacity}
\end{figure}

@@

The critical core mass declines with distance from the star, as shown in \Fig{fig:Mcvsa} for our standard disk model.   The main reason for the decline, as explained above, is that atmospheres grow faster at lower temperatures.  The lower densities and pressures in the outer disk have a much smaller effect.  Since the vast majority of disk models have a disk temperature that declines with $a$, our qualitative result is robust.   We also show that higher $\mu$ values give lower values of $M\crit$.

The average power-law decline from 1 to 100 AU (not plotted) is $M\crit \propto a^{-0.3}$, for both choices of $\mu$.   While not a drastic decline, the ability of distant low mass cores to accrete gas efficiently is significant for the interpretation of direct imaging surveys.  However, our model offers no guarantee of copious giant planets at large distances.  Many histories of solid accretion are possible, and solid cores may grow too slowly to allow the rapid gas accretion that we model.

Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $M\crit$ with $a$.  This discrepancy is not surprising, as we have shown that self-gravity strongly affects evolution, even before runaway growth begins and the crossover mass is reached.

\Fig{fig:Mcritopacity} shows that reducing the opacity by an order of magnitude significantly reduces $M\crit$.  Furthermore, this opacity effect is stronger at larger distances.  The opacity reduction by a factor of 10 lowers $M\crit$ at 5 AU by a factor of $\sim$$2.5$ and at 100 AU by a factor of $\sim$$3.5$.  Atmospheric opacity is a dominant uncertainty in core accretion modelling.  However, unless atmospheric opacity varies significantly with disk radius, the general decline in $M\crit$ with increasing $a$ should hold.  For our scenario of negligible ongoing planetesimal accretion, it is tempting to think that dust could settle out of the radiative zone, lowering the opacity \citep{podolak03}.  Nonetheless, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.

@@

\begin{enumerate}
\item The minimum or critical core mass, $M\crit$, for giant planet formation declines with stellocentric distance in standard protoplanetary disk models.  For our reference case, the critical mass is $\sim$$8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim$$3.5 M_{\oplus}$ at $a =100$ AU.  This decline roughly follows $M\crit \propto a^{-0.3}$.

@@

\item A larger mean molecular weight reduces critical core masses, in agreement with \citet{HorIko11}.  If enrichment in heavy elements correlates with increased dust opacity, then the stronger opacity effect will dominate, increasing $M\crit$.

@@

Thus for giant planets to form around cores near $M\crit$, core growth must stall quite severely.   The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.  On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.  In these cases, relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.   Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.

@@

\subsection{The Opacity Effect}
\label{opacityan}
A  lower opacity  decreases the critical core mass.  Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, specifically  $M_{\rm{crit}} \approx9 M_\oplus$ for the parameters in \Eq{eq:Mcrit}.  The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.

>>>>

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_new.eps}
%\vspace{-0.5in}
\caption{The critical core mass as a function of semimajor axis, for a disk lifetime of $3$ Myrs and two values of the mean molecular weight ($\mu = 2.35$ is for Solar abundances). The decline in $\MC$ with distance is a robust result for standard disk models.  The analytic model, which neglects self-gravity, over-predicts the steepness of the decline.}
\label{fig:Mcvsa}
\end{figure}

\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Mcrit_vs_a_3Myrs_opacity.eps}
%\vspace{-0.5in}
\caption{Critical core masses vs. distance for standard and reduced (by a factor of ten) dust opacities.  Lower opacities give significantly lower $\MC$ values.  Atmospheric opacity remains a large uncertainty in core accretion models.}
\label{fig:Mcritopacity}
\end{figure}

@@

The critical core mass declines with distance from the star, as shown in \Fig{fig:Mcvsa} for our standard disk model.   The main reason for the decline, as explained above, is that atmospheres grow faster at lower temperatures.  The lower densities and pressures in the outer disk have a much smaller effect.  Since the vast majority of disk models have a disk temperature that declines with $a$, our qualitative result is robust.   We also show that higher $\mu$ values give lower values of $\MC$.

The average power-law decline from 1 to 100 AU (not plotted) is $\MC \propto a^{-0.3}$, for both choices of $\mu$.   While not a drastic decline, the ability of distant low mass cores to accrete gas efficiently is significant for the interpretation of direct imaging surveys.  However, our model offers no guarantee of copious giant planets at large distances.  Many histories of solid accretion are possible, and solid cores may grow too slowly to allow the rapid gas accretion that we model.

Our non-self-gravitating analytic model (dashed curves in \Fig{fig:Mcvsa}) over-predicts the steepness of the decline in $\MC$ with $a$.  This discrepancy is not surprising, as we have shown that self-gravity strongly affects evolution, even before runaway growth begins and the crossover mass is reached.

\Fig{fig:Mcritopacity} shows that reducing the opacity by an order of magnitude significantly reduces $\MC$.  Furthermore, this opacity effect is stronger at larger distances.  The opacity reduction by a factor of 10 lowers $\MC$ at 5 AU by a factor of $\sim$$2.5$ and at 100 AU by a factor of $\sim$$3.5$.  Atmospheric opacity is a dominant uncertainty in core accretion modelling.  However, unless atmospheric opacity varies significantly with disk radius, the general decline in $\MC$ with increasing $a$ should hold.  For our scenario of negligible ongoing planetesimal accretion, it is tempting to think that dust could settle out of the radiative zone, lowering the opacity \citep{podolak03}.  Nonetheless, since opacity near the RCB is a crucial factor, we speculate that convective overshoot may prevent very low opacities.

@@

\begin{enumerate}
\item The minimum or critical core mass, $\MC$, for giant planet formation declines with stellocentric distance in standard protoplanetary disk models.  For our reference case, the critical mass is $\sim$$8.5 M_{\oplus}$ at $a = 5$ AU, decreasing to $\sim$$3.5 M_{\oplus}$ at $a =100$ AU.  This decline roughly follows $\MC \propto a^{-0.3}$.

@@

\item A larger mean molecular weight reduces critical core masses, in agreement with \citet{HorIko11}.  If enrichment in heavy elements correlates with increased dust opacity, then the stronger opacity effect will dominate, increasing $\MC$.

@@

Thus for giant planets to form around cores near $\MC$, core growth must stall quite severely.   The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.  On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk.  In these cases, relatively high mass cores would remain as ice giants, akin to Uranus and Neptune.   Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.

@@

\subsection{The Opacity Effect}
\label{opacityan}
A  lower opacity  decreases the critical core mass.  Reducing the opacity by a factor of one hundred results in a critical core mass one order of magnitude lower than in the standard ISM case, specifically  $\MC \approx9 M_\oplus$ for the parameters in \Eq{eq:Mcrit}.  The reduction is not  as strong as the nominal scaling would imply, $0.01^{3/5} \approx 0.06$, because $\xi$ increases.
